---
title: "Practical 2"
author: "Jonathan Bartlett"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

# Introduction
In this practical you will perform analyses of a real dataset to explore and develop your understanding of missing data mechanisms, the conditions under which complete case analyses are valid, and multiple imputation of a single variable with missing values.

The practical will use data from the [1999-2000 National Health and Nutrition Examination Survey (NHANES)](https://wwwn.cdc.gov/nchs/nhanes/ContinuousNhanes/Default.aspx?BeginYear=1999) conducted in the the USA. The dataset we will use contains a small subset of the many variables collected at this survey. It has also been linked to [mortality data](https://www.cdc.gov/nchs/data-linkage/mortality-public.htm), so that we have follow-up for mortality from the survey at 1999-2000 through to around 2015. The survey was conducted using complex survey sampling techniques, which for the purposes of these illustrative practicals we will ignore.

Load the dataset and summarize its contents. Click on the `nhanesMort` data frame in the environment viewer to look at the dataset and the variable descriptions.
```{r, results='hide'}	
load("../Datasets/nhanesMort.Rdata")
summary(nhanesMort)
```

The data summary shows that quite a few of the variables collected at the survey have some missing data. There are no missing values in the mortality status variable `dead` nor the corresponding time to death/censoring, `tMonths`.

# Alcohol consumption
The first analysis of the NHANES data we will consider is a marginal analysis of the alcohol variable `ALQ150` we have, which is the respondent's answer to the question 'Was there ever a time or times in your life when you drank 5 or more drinks of any kind of alcoholic beverage almost every day?'. Since this variable is binary (yes/no), this analysis aims to simply estimate the proportion whose answer to this question is yes. The data summary earlier showed that of the `r dim(nhanesMort)[1]` observations in the dataset, `r sum(is.na(nhanesMort$ALQ150))` are missing this variable.

## Complete case analysis
First perform a complete case analysis:
```{r, eval=FALSE}
prop.table(table(nhanesMort$ALQ150))
```
For comparison later on, we save this result:
```{r, eval=FALSE}
ccaALQYes <- prop.table(table(nhanesMort$ALQ150))[2]
```

**Next, perform a logistic regression analysis to see whether and how missingness in `ALQ150` is associated with age, ethnicity, and gender. What do you conclude about the missingness mechanism for `ALQ150`?**

```{r, echo=FALSE, eval=FALSE}
summary(glm(is.na(ALQ150)~age+ethnicity+gender, data=nhanesMort, family="binomial"))
```

To further help judge whether the complete case analysis is likely to be biased, and if so, how, perform an analysis to investigate how the probability of saying Yes to the `ALQ150` question is associated with the covariates age, ethnicity, gender and weight.

```{r, echo=FALSE, eval=FALSE}
summary(glm(ALQ150~age+ethnicity+gender, data=nhanesMort, family="binomial"))
```

**Given the results of your analyses, do you think the complete case proportion of Yes estimated earlier is unbiased, biased upwards, or biased downwards?**

## Multiple imputation
We will now use multiple imputation to impute the missing `ALQ150` values assuming they are MAR. In this practical we will impute `ALQ150` using the other fully observed baseline variables age, ethnicity and gender. In the next practical we will see how we can exploit the other partially observed variables in the imputation process.

We will create the multiple imputations using the [mice](https://cran.r-project.org/package=mice) package, so we must first install it if you haven't done so already, and then load it:
```{r, eval=FALSE}
#install.packages("mice")
library(mice)
```

The `mice` function performs imputation. It will attempt to impute all missing values in the data frame you pass it. Since we will cover the approach it uses for this in the next session, here we will first create a subset of the data which only contains the variable we want to impute, `ALQ150`, and the fully observed variables `age`, `ethnicity` and `gender` that we will use as covariates in the imputation model.
```{r, eval=FALSE}
nhanes_subset <- subset(nhanesMort, select=c("age", "ethnicity", "gender", "ALQ150"))
```

Now we are ready to impute the missing values in `ALQ150`. MI involves drawing random numbers. Before we impute, we set R's random number seed to some value. The value we choose is not important, but setting it to the same value means that our results are re-producible and that we should all obtain the same 'random' values. 

```{r, eval=FALSE}
set.seed(7341)
imps <- mice(nhanes_subset, m=10, maxit=1)
```

We ask to `mice` to create `m=10` imputed datasets. We will discuss the `maxit` argument in the next session. After the imputation has completed, we can summarise the resulting object:

```{r, eval=FALSE}
summary(imps)
```

Now we can analyse the imputed datasets. The `mice` package contains functions which will apply Rubin's rules combination rules for us. However they require the function that performs the full data analysis to return estimated coefficients and standard errors. To do this, rather than use `prop.table(table(nhanesMort$ALQ150))` as we did before, we will fit a logistic regression model for `ALQ150` without any covariates. The only parameter in this model is an intercept, and its value corresponds to the log odds of answering Yes to `ALQ150`:

```{r, eval=FALSE}
fit <- with(imps, exp=glm(ALQ150~1, family="binomial"))
summary(pool(fit), conf.int = TRUE)
```

In order to compare to the estimate we obtain earlier, we now transform the estimated log odds of answering Yes from the MI analysis to a probability/proportion, and compare it to the complete case estimate we obtained earlier:

```{r, eval=FALSE}
as.numeric(exp(summary(pool(fit))[1])/(1+exp(summary(pool(fit))[1])))
#compared to our earlier CCA estimate
ccaALQYes
```

**Does the change in estimate agree with what you predicted earlier?**

**Under what assumption is your MI analysis valid? Do you think this assumption is plausible, and more plausible than the assumption required for the CCA estimate to be valid?**

# Analysis of mortality
If you have time, you can consider a second substantive analysis, which we will also look at in the next practical. Our second substantive analysis will be a model for how the probability of death (all-causes) within 10 years relates to the various 'baseline' variables which were measured at the year 1999-2000 survey.

First we will use the event indicator `dead` and the time variable `tMonths` to construct our binary outcome variable. Before we do this, we look at whether any individuals were censored for mortality before 10 years:
  
```{r, eval=FALSE}
summary(nhanesMort$tMonths[nhanesMort$dead==0])
```

This shows that the first censoring took place at 15 years follow-up. We can therefore proceed to create a binary variable for 10 year mortality without any problems with censoring:
  
```{r, eval=FALSE}
nhanesMort$dead10 <- 1*((nhanesMort$dead==1) & (nhanesMort$tMonths<120)) 
```

Next fit a complete case logistic regression model for the `dead10` variable, with `gender`, `age`, `ethnicity`, `sbp`, `waist_circum`, `weight`, `total_cholesterol_chol`, `hdl` and `ALQ150` as covariates:

```{r, eval=FALSE}
cca10year <- glm(dead10~gender+age+ethnicity+sbp+waist_circum+weight+total_chol+hdl+ALQ150, data=nhanesMort, family="binomial")
summary(cca10year)
```

**How many observations have been omitted due to missing data? What is the impact of this on the estimates, standard errors and confidence intervals?**
  
**Under what conditions would this complete case analysis be unbiased? Do you think these conditions are plausible here? Perform some data analyses to investigate. You may find the `complete.cases()` function useful.**

```{r, eval=FALSE, echo=FALSE}
nhanesMort$cc <- complete.cases(nhanesMort)
summary(glm(cc~dead10, data=nhanesMort, family="binomial"))
#add in fully observed covariates
summary(glm(cc~dead10+gender+age+ethnicity, data=nhanesMort, family="binomial"))
```

In the next practical we will look at how MI can be applied to impute the missing values in all the variables.